---
title: "Tidy Data Lite: Exercises"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, error = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(dataAnim)
library(DSR)
```

## pivot_longer: One variable spread across multiple columns

- Column headers are values, not variables

**The pivot_longer function can help!**

```{r}
datoy_wide
```

- 3 variables
  + `Name` stored in rows
  + `Subject` spread across column names
  + `Score` stored in cell values
  
To tidy we need to use pivot_longer to make the following transformation

[pivot longer animation](./pivot_longer_animation.html)

- Essential arguments
  + `data`specifies the df to reshape `datoy_wide`
  + `cols` specifies the columns that needs to be reshaped, in this case it's `English & Maths`. Hint: the columns that contain values, not variables, as headers
  + `names_to` a string specifying the name of the new column to create from the data stored in the headers of the columns specified in the `cols` argument, i.e. `Subject`
  + `values_to` a string specifiying the name of the new column to create from the data data stored in the cell values of the columns specified in the `cols` argument, i.e. `Score`

Let's code it!

```{r}
?pivot_longer

#datoy_wide  %>% 
#  pivot_longer(
#  )
```

## You try: Religion and Income!

Uncomment the code below and fill in to tidy the relig_income dataset 

```{r}

relig_income %>% 
  head()

#relig_income %>% 
# pivot_longer(
#   cols = , 
#   names_to = "income",
#   values_to = "count"
#   )

```

Let's try again. 

Uncomment the code below and fill in to tidy the relig_income dataset

- Try specifying the columns to pivot into a longer format in the `cols` argument in a different way than you did above. 
  + Remember we can use tidyselect principles to specify columns
  + Be sure to fill in arguments `names_to` and `values_to`

```{r}
relig_income %>% 
  head()

#relig_income %>% 
# pivot_longer(
#   cols = , 
#   names_to = ,
#   values_to = 
#   )

```

<!-- Which of the following would NOT have worked? -->

<!-- 1. pivot_longer(data = relig_income,cols = -c(religion), names_to = "income", values_to = "count") -->

<!-- 2. pivot_longer(data = relig_income,cols = 2:11, names_to = "income", values_to = "count") -->

<!-- 3. pivot_longer(data = relig_income,cols = `<$10k`:`Don't know/refused`, names_to = "income", values_to = "count") -->

<!-- 4. pivot_longer(data = relig_income,cols = <$10k:Don't know/refused, names_to = "income", values_to = "count") -->

<!-- >syntactically valid names consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number -->

## pivot_wider: An observation is scattered across multiple rows

- Some values are actually variables (i.e., variables stored in a column)
  + In this instance, we would use pivot_wider to make tidy data
  
```{r, echo = FALSE}
weather <- readr::read_csv("weather.csv") %>% 
  pivot_longer(d1:d31, names_to = "day", names_prefix = "d",
               values_to = "value", values_drop_na = TRUE) 
weather %>% 
  head()

```

- Most variables have individuals columns (`id`, `year`, `month`, `day`, `value`) but two variables
(`tmax`, `tmin`) are spread across rows, and stored in the `element` column.
  + `element` is not a variable - it's a column that stores the names of two variables
  
- Fixing this requires `pivot_wider` function 
  + Performs the inverse of `pivot_longer` by spreading the `element` and `value` columns back out into columns
  
```{r}
weather %>% 
  pivot_wider(names_from = element, values_from = value) %>% 
  head()

?pivot_wider
```

- Essential Arguments
  + `data`specifies the df to reshape `weather`
  + `names_from` and `values_from` are paired arguments describing which column(s) R should get the name of the output column(s) from, `name_from` = `element`, and which column(s) it should get the cell values from `values_from` = `value`
    - `names_from` usually column(s) that contains the names of variables as values
    - `values_from` the column(s) whose values correspond to the variable(s) stored in the `names_from` columns(s) 

- In practice, relatively rare to need pivot_wider for tidying, but it's often useful for creating summary tables for presentation

Remember we tidied the scores data 

```{r}
datoy_long

```

What if we wanted this in wide format for a summary table?

[pivot wider animation](./pivot_wider_animation.html)


Let's code it!

```{r}
datoy_long 
```


## You try!

Take our "tidied" relig_income data set and put it back in its original "wide" format by uncommenting the code below

```{r}


tidy_relig_income <- relig_income %>% 
  pivot_longer(cols = -religion, names_to = "income", values_to = "count")

# have
tidy_relig_income

#want
relig_income

# tidy_relig_income %>% 
#   pivot_wider(
#     names_from = ,
#     values_from = 
#   )

```

## Putting it all together

Load & examine the following dataset 

```{r}
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

stocks
```

Make a new dataset "tidy_stocks" by using the appropriate pivot function to tidy. 
  - replace "???" with longer or wider

```{r}
# insert tidying code here

# tidy_stocks <- stocks %>% 
#   pivot_???()
```


## The bells and whistles: for the big messes!

- pivot_longer:

  + Numeric data in column names
  + Many variables in column names
  + Multiple observations per row
  + Duplicate column names
  + Columns and rows are values of variables

- pivot_wider: 

  + values from multiple columns
  + filling in missing values
  + for aggregation/summary table creation
  + column names from multiple variables
  + multi-choice data and id_cols

- pivot_longer + pivot_wider


- We will dive in after break!

## How am I doing?

Content | Presentation
------- | ------------
   +    |     +
   -    |     -
   
