---
title: "Tidy Data Lite: Exercises"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, error = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(dataAnim)
library(DSR)
```

## pivot_longer: One variable spread across multiple columns

- Column headers are values, not variables

**The pivot_longer function can help!**

```{r}
datoy_wide
```

- 3 variables
  + *Name* stored in rows
  + *Subject* spread across column names
  + *Score* stored in cell values
  
To tidy we need to use pivot_longer to make the following transformation

[pivot longer animation](./pivot_longer_animation.html)

- The first argument is the *data* to reshape *datoy_wide*
- The second argument describes which *cols* needs to be reshaped, in this case it's *English & Maths*
- The *names_to* argument gives the name of the variable that will be created from the data stored in the column names, i.e. *Subject*
- The *values_to* gives the name of the variable that will be created from the data stored in the cell value, i.e. *Score*
    + Neither the names_to nor the values_to column exists in datoy_wide, so we provide them as character strings surrounded in quotes.

Let's code it!

```{r}
?pivot_longer

#datoy_wide  %>% 
#  pivot_longer(
#  )
```

## You try: Religion and Income!

Uncomment the code below and fill in to tidy the relig_income dataset 

```{r}

relig_income %>% 
  head()

#relig_income %>% 
# pivot_longer(
#   cols = , 
#   names_to = "income",
#   values_to = "count"
#   )

```

Let's try again. 

Uncomment the code below and fill in to tidy the relig_income dataset

- Try a different way of selecting "cols" than we did above (i.e., -religion) and be sure to fill in arguments names_to and values_to

```{r}
relig_income %>% 
  head()

#relig_income %>% 
# pivot_longer(
#   cols = , 
#   names_to = ,
#   values_to = 
#   )

```

Which of the following would NOT have worked?

1. pivot_longer(data = relig_income,cols = -c(religion), names_to = "income", values_to = "count")

2. pivot_longer(data = relig_income,cols = 2:11, names_to = "income", values_to = "count")

3. pivot_longer(data = relig_income,cols = `<$10k`:`Don't know/refused`, names_to = "income", values_to = "count")

4. pivot_longer(data = relig_income,cols = <$10k:Don't know/refused, names_to = "income", values_to = "count")
 
>syntactically valid names consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number

## pivot_wider: An observation is scattered across multiple rows

- Some values are actually variables (i.e., variables stored in a column)
  + In this instance, we would use pivot_wider to make tidy data
  
```{r, echo = FALSE}
weather <- readr::read_csv("weather.csv") %>% 
  pivot_longer(d1:d31, names_to = "day", names_prefix = "d",
               values_to = "value", values_drop_na = TRUE) 
weather %>% 
  head()

```

- Most variables have individuals columns *(id, year,month, day, value)* but two variables
*(tmax, tmin)* are spread across rows, and stored in the "element" column.
  + Element is not a variable - it's a column that stores the names of two variables
  
- Fixing this requires **pivot_wider** function 
  + Performs the inverse of **pivot_longer** by spreading the *element* and *value* columns back out into columns
  
```{r}
weather %>% 
  pivot_wider(names_from = element, values_from = value) %>% 
  head()

?pivot_wider
```

- The first argument is the *data* to reshape *weather*
- The second argument,*id_cols* describes which columns that unique identify each observation. The default is usually sufficient (defaults to all columns in the data except for the columns specified in names_from and values_from)
  + Typically only used when you have an additional variable that is directly related. Can be helpful in cleaning multiple choice data
- The *names_from* and *values_from* are paired arguments describing which column (or columns) to get the name of the output column *(name_from = element)*, and which column (or columns) to get the cell values from *(values_from = value)*

- In practice, relatively rare to need pivot_wider for tidying, but it's often useful for creating summary tables for presentation

Remember we tidied the scores data 

```{r}
datoy_long

```

What if we wanted this in wide format for a summary table?

[pivot wider animation](./pivot_wider_animation.html)


Let's code it!

```{r}
datoy_long 
```


## You try!

Take our "tidied" relig_income data set and put it back in its original "wide" format by uncommenting the code below

```{r}


tidy_relig_income <- relig_income %>% 
  pivot_longer(cols = -religion, names_to = "income", values_to = "count")

# have
tidy_relig_income

#want
relig_income

# tidy_relig_income %>% 
#   pivot_wider(
#     names_from = ,
#     values_from = 
#   )

```

<!-- One More -->

<!-- Tidy "table2" using pivot_wider -->

<!-- HINT: -->

<!-- - This dataset is actually untidy, like the weather example, in that some values are actually variables (i.e., variables stored in a column) -->


<!-- ```{r} -->
<!-- table2 %>%  -->
<!--   head() -->

<!-- # insert tidying code here -->

<!-- # table2 %>%  -->
<!-- #   pivot_wider( -->
<!-- #   ) -->
<!-- ``` -->


## Putting it all together

Load & examine the following dataset 

```{r}
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

stocks
```

Make a new dataset "tidy_stocks" by using the appropriate pivot function to tidy

```{r}
# insert tidying code here

# tidy_stocks <- stocks %>% 
#   pivot_???()
```


## When things are a real mess!

- pivot_longer:

  + Numeric data in column names
  + Many variables in column names
  + Multiple observations per row
  + Duplicate column names
  + Columns and rows are values of variables

- pivot_wider: 

  + values from multiple columns
  + filling in missing values
  + for aggregation/summary table creation
  + column names from multiple variables
  + multi-choice data and id_cols

- pivot_longer + pivot_wider


- We will dive in after break!

## How am I doing?

Content | Presentation
------- | ------------
   +    |     +
   -    |     -
   
